---
title: "TLS_score"
output: html_document
date: "2024-05-06"
---

The patients will be classified into different groups first using TLS signatures genes and TLS-related cells (MP, T, B, DC).


```{r library}

library(Seurat)
#library(leiden)
#library(leidenAlg)
library(ggplot2)
#library(sctransform)
#library(scDblFinder)
#library(SingleCellExperiment)
#library("viridis")
#library(knitr)
library(dplyr)
#library(readxl)
#library(SingleR)
#library(celldex)
library(ggplot2)
library(patchwork)
#library ("harmony")
library (hdf5r)

```
# Loading objects 

```{r}
GSE_merge = readRDS("/omics/groups/OE0436/internal/Linh/Datasets/Merged/GSE_merge_annotated") #GSE 3 datasets,annotated
saveRDS (patient_list, "/omics/groups/OE0436/data/Linh/Datasets/DEA/patient_list") #only immune cells, around 25k. Saved 13.05 - run second time


```

# Importing datasets

```{r Import core signatures}
tls_core_signature <- c("CCL2","CCL3","CCL4","CCL5","CCL8","CCL18","CCL19","CCL21","CXCL9","CXCL10","CXCL11","CXCL12", "CXCL13","CD79B", 
"CETP","LAT", "CD1D", "PTGDS","CXCR5","SELL","ICOS","SH2D1A","TIGIT","PDCD1",                 
"BANK1","CD22","CD79A","CR1","FCRL2", "MS4A1", "FCER2", "LAMP3","CD86","CD80","CD83","CCR7")

```

```{r Import GSE}

GSE_all_sub = readRDS("/omics/groups/OE0436/data/Linh/Datasets/Merged/GSE_all_sub") #object no scale. 4 datasets
GSE186344 = subset (GSE_all_sub, subset = dataset_ID == "GSE186344")
GSE174401 = subset (GSE_all_sub, subset = dataset_ID == "GSE174401")
GSE234832 = subset (GSE_all_sub, subset = dataset_ID == "GSE234832")
GSE_merge <- merge(x = GSE186344, y = c(GSE174401,GSE234832))

features = Features (GSE_merge) # extracting features names . 34k genes
cells = Cells (GSE_merge) # extracting cell names . 96k cells 
```

```{r Normalize the data}
GSE_merge@assays[["RNA"]] = JoinLayers(GSE_merge@assays[["RNA"]]) #merge layers
GSE_merge@assays[["RNA"]] = NormalizeData(GSE_merge@assays[["RNA"]]) #default using Lognormlize, column wise 
```

# Adding cell type annotation and subset cell types

```{r}

#ct_ann_GSE344_10kSingleR = readRDS ("/omics/groups/OE0436/data/Linh/Datasets/Annotation/ct_ann_GSE344_10kSingleR")
GSE_merge@meta.data[["ct_ann_GSE344_10kSingleR"]] = ct_ann_GSE344_10kSingleR@listData[["pruned.labels"]] #adding the metadata info 

saveRDS(GSE_merge,"/omics/groups/OE0436/data/Linh/Datasets/Merged/GSE_merge_annotated") #GSE 3 datasets,annotated

# subsetting for a group of cells for classifcation 
cell_sub = c("MAMs:APOE+", "MAMs:S100A8+","T:CD4+:CM1", "T:CD8+:EM","Treg", "T:CD4+:CM2","T:CM", "B-c2","B-c1", "cDC2:CD1C+/CLEC10A+")

GSE_list = list()
GSE_sub = subset (GSE_merge, subset = ct_ann_GSE344_10kSingleR == "MAMs:APOE+")  #initalize GSE_sub, otherwise wont work
for (i in 2:length(cell_sub)) {
  GSE_list [[i]] = subset (GSE_merge, subset = ct_ann_GSE344_10kSingleR == cell_sub[i])  # Warning: Removing 498 cells missing data for vars requested Warning: longer object length is not a multiple of shorter object length
  GSE_sub = merge(GSE_sub, GSE_list[[i]])
}

#rm (GSE_list) # to free memory
# Join layers
GSE_sub@assays[["RNA"]] = JoinLayer(GSE_sub@assays[["RNA"]]) #merge layers

# number of cells 
length(Cells(GSE_sub)) #25900 vs 100k cells (30%)

```

# Creating list of patients
 
At the end I get a list of patients, each element is a sparse matrix contains normalised data for TLS signature for a patient

```{r }
patients = unique (GSE_sub@meta.data[["sample_ID"]])
patient_list = list ()

for (i in 1:length(patients)){
  seurat_patient = subset (GSE_sub, subset = sample_ID == patients [i]) 
  matrix_patient = seurat_patient@assays[["RNA"]]@layers[["data"]] # extracting dgcmatrix (matrix patient contain norm data for each patient)
  matrix_patient@Dimnames[[1]] = Features(seurat_patient)  # give genes name
  matrix_patient@Dimnames[[2]] = Cells(seurat_patient) # give cell barcode
  matrix_patient  = matrix_patient [tls_core_signature, ] # subsetting for TLS signatures
  
  patient_list[[patients[i]]] = matrix_patient  #adding to a list
}
saveRDS (patient_list, "/omics/groups/OE0436/data/Linh/Datasets/DEA/patient_list") 
patient_list = readRDS ("/omics/groups/OE0436/data/Linh/Datasets/DEA/patient_list")
```

# Calculating mean + normalised for each patients

```{r}

for (i in 1:length(patient_list)) {
  mean = rowMeans (patient_list [[i]]) #calculating the mean
  no_cells = length(patient_list[[i]]@Dimnames[[2]])
  patient_list [[i]] = mean/no_cells # normalised the value by the number of cell for that patient
}
### no diving again

for (i in 1:length(patient_list)) {
  mean = rowMeans (patient_list [[i]]) #calculating the mean
  patient_list [[i]] = mean
}
```

# Merge all patients into one dataframe 
```{r}
patient_df = do.call(cbind,lapply(patient_list, as.data.frame)) # do.call(): call a function with a list of arguments. Useful when: have a list of arguments (stored as a list) and want to pass a func to it. lapply () take 2 arguments: the list or vector to apply func to and the func to apply. Short for list apply. lapply and sapply: lapply always return a list
colnames(patient_df) = patients
```


# Calculating TLS score 
calculating the geometric mean value as TLS score

```{r}

library(psych)
patient_df_try = patient_df*100
patient_df_try = patient_df_try + 1 #to handle 0 values

geometric_means <- apply(patient_df_try, 2, geometric.mean) #2: columns # alot of 0 initially 

geometric_means 

```

```{r}
#Box Plot based on Location and Primary Origin
# extracting the patients ID 
patient_ID = names(geometric_means)
geometric_means = as.data.frame(geometric_means)
geometric_means = cbind (geometric_means, rownames(geometric_means))
colnames(geometric_means) = c("score", "patients")

score_only = geometric_means$score
geometric_means$score = as.numeric(geometric_means$score)
#boxplot (as.numeric(score_only)) 
#points(as.numeric(score_only))

png ("/omics/groups/OE0436/data/Linh/Output/DEA/TLS_Score_subset.png",width = 700, height = 1000)
ggplot(geometric_means, aes(x = "Patient_ID",y = as.numeric(score))) + geom_boxplot(fill = "transparent") + geom_point(color="blue", size= 2 , alpha=0.9) + labs(titel="TLS Signature Scores", y = "Scores")
dev.off()

```

# 30k cells 
geometric_means 
GSE186344_888 GSE186344_889 GSE186344_890 GSE186344_891 GSE186344_892 GSE186344_893 GSE186344_895 GSE186344_896 
    11.189148     43.030443      6.266978      5.397498      4.018999      2.747358      4.753188      2.708639 
GSE186344_897 GSE186344_898 GSE186344_900 GSE186344_902 GSE186344_904 GSE174401_902 GSE174401_904 GSE174401_905 
    44.619893      4.228860      3.844892     38.199140      4.642268      5.362595     10.869848     14.098419 
GSE174401_906 GSE174401_907 GSE174401_908 GSE174401_909 GSE174401_910 GSE174401_911 GSE174401_912 GSE174401_913 
     7.240936      6.675551     32.777981     11.545989     26.202997      4.077531      6.394087     12.852366 
GSE174401_914 GSE174401_915 GSE234832_324 GSE234832_325 GSE234832_326 GSE234832_327 GSE234832_328 GSE186344_894 
    12.695632      5.611733      8.717353      8.938101     13.702151      7.215077     12.218112      7.250461 
GSE174401_903 
     6.952759 
     
# 100k cells     
GSE186344_888 GSE186344_889 GSE186344_890 GSE186344_891 GSE186344_892 GSE186344_893 GSE186344_895 GSE186344_896 
     7.526728      1.133938      3.124993      1.768977      1.259755      1.034044      1.078502      2.824103 
GSE186344_897 GSE186344_898 GSE186344_900 GSE186344_902 GSE186344_904 GSE174401_902 GSE174401_904 GSE174401_905 
     2.719638      1.331635      1.815544      2.615125      6.698268      1.872873      1.735602      1.464912 
GSE174401_906 GSE174401_907 GSE174401_908 GSE174401_909 GSE174401_910 GSE174401_911 GSE174401_912 GSE174401_913 
     2.614255      6.457173      7.599788      5.297701      1.826113      5.274608      1.151241      3.884681 
GSE174401_914 GSE174401_915 GSE234832_324 GSE234832_325 GSE234832_326 GSE234832_327 GSE234832_328 GSE186344_894 
     6.089908      1.437819     11.109638      4.223080      5.500000      6.289674      9.254261      3.491245 
GSE174401_903 
     6.376086 
     
# 30k cells, not divide by cell number again. (different scale, * by 1000 and not 10000)
GSE186344_888 GSE186344_889 GSE186344_890 GSE186344_891 GSE186344_892 GSE186344_893 GSE186344_894 GSE186344_895 
    14.209667      6.003466      8.362432      6.322837      4.043580      2.011474      2.352718      5.908778 
GSE186344_896 GSE186344_897 GSE186344_898 GSE186344_900 GSE186344_902 GSE186344_904 GSE174401_902 GSE174401_903 
     8.059223      5.465833      6.174158      4.031106      9.144434      6.097455      2.128250      4.016213 
GSE174401_904 GSE174401_905 GSE174401_906 GSE174401_907 GSE174401_908 GSE174401_909 GSE174401_910 GSE174401_911 
     5.179223      4.503614      7.582296      6.022782     11.094672      3.856954      6.135402      7.604693 
GSE174401_912 GSE174401_913 GSE174401_914 GSE174401_915 GSE234832_324 GSE234832_325 GSE234832_326 GSE234832_327 
     3.471459      4.898340      7.143851      6.645914      6.398503      6.312162      5.238519      5.499491 
GSE234832_328 
     2.320702 
     
#cell distribution in the patients 
```{r}
############### create a list for all patients, each contains the cell type for all cells for that patient 
patients = unique(GSE_merge@meta.data[["sample_ID"]]) 
patients_list = list ()

patients_list <- lapply(patients, function(patient) which(GSE_merge@meta.data[["sample_ID"]] == patient)) #create a list in which each element is the position of cells of that patient in seurat obj
patients_list = lapply (patients_list, function(patient_inlist) GSE_merge@meta.data[["ct_ann_GSE344_10kSingleR"]] [patient_inlist])  #each element: cell type of that patients 

patients_list <- setNames(patients_list, patients) # names
#MP_count = sum(patients_list[[1]] %in% MP)
################################## create a df that counts the cell type for each patients. column: patients ID, row: MP, T, B, DC, MTCs, all immune, all cell, cancer type
celltype = data.frame()
MP = c("MAMs:S100A8+","MAMs:APOE+")
T_cells = c("T:CD4+:CM1", "T:CD8+:EM","Treg", "T:CD4+:CM2","T:CM")
B = c("B-c2","B-c1")
DC = "cDC2:CD1C+/CLEC10A+" 
MTC = c("MTCs", "MTC")

celltype <- lapply(seq_along(patients_list), function(i) {
  c(
    MP_count = sum(patients_list[[i]] %in% MP, na.rm = TRUE),
    T_cells_count = sum(patients_list[[i]] %in% T_cells),
    B_count = sum(patients_list[[i]] %in% B, na.rm = TRUE),
    DC_count = sum(patients_list[[i]] %in% DC, na.rm = TRUE),
    MTC_count = sum(patients_list[[i]] %in% MTC, na.rm = TRUE)
  )
})

celltype <- as.data.frame(do.call(rbind, celltype))
#celltype$all_immune <- rowSums(celltype[, c("MP_count", "T_cells_count", "B_count", "DC_count")], na.rm = TRUE)
for (i in 1:33) {
  celltype$others [i] = length (patients_list[[i]]) - rowSums(celltype[i, c("MP_count", "T_cells_count", "B_count", "DC_count", "MTC_count")], na.rm = TRUE)
}

rownames (celltype) = names (patients_list) 

saveRDS(celltype, "/omics/groups/OE0436/data/Linh/Datasets/DEA/cell_type")

ID = lapply(patients, function(patient) which(GSE_merge@meta.data[["sample_ID"]] == patient))

celltype$cancer_type  = lapply(ID, function(ID_pat) unique(GSE_merge@meta.data[["cancer_types"]][ID_pat]))

celltype$TLS_Score  = geometric_means

```


```{r}
# Stacked
library(ggplot2)

# Reorder the rows of 'celltype' to match the order of patients you want on the x-axis
# If 'patients' is a vector containing patient IDs, you can reorder based on that
celltype <- celltype[match(patients, rownames(celltype)), ]

# Convert 'celltype' dataframe to long format for ggplot
library(tidyr)
patient_IDs = rownames(celltype)

# include all the INFOS
All_infos = c()
All_infos <- c(sapply(strsplit(celltype$patient_IDs, "_"), function(x) x[[2]]))
All_infos = paste (All_infos,celltype$cancer_type, sep = "_")
All_infos = paste (All_infos,celltype$cancer_type, sep = "_")
All_infos = paste (All_infos,as.character(round(celltype$TLS_Score)), sep = "_")

## creating format to plot stacked barplot

celltype = cbind (All_infos,celltype)
celltype_long <- pivot_longer(celltype, cols = c(MP_count, T_cells_count, B_count, DC_count, MTC_count, others), names_to = "Cell_Type", values_to = "Count")
celltype_longtry = celltype_long [,c("All_infos","Cell_Type","Count")]
 

# Stacked
png ("/omics/groups/OE0436/data/Linh/Output/DEA/StackedBarPlotofCellCounts.png",width =1000, height = 700)
ggplot(celltype_longtry, aes(fill=Cell_Type, y=Count, x=All_infos)) + 
    geom_bar(position="stack", stat="identity")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Stacked Bar Plot of Cell Counts for Each Patient")
dev.off()

```

```{r Bar plots for other cells as well}
# MP
sample_name = rownames (celltype)
celltype = cbind(sample_name,celltype)

celltype_long <- pivot_longer(celltype, cols = c(MP_count), names_to = "Cell_Type", values_to = "Count")
celltype_longtry = celltype_long [,c("sample_name","Cell_Type","Count")]
# Stacked
png ("/omics/groups/OE0436/internal/Linh/Output/DEA/MP_Count.png",width =700, height = 550)
MP = ggplot(celltype_longtry, aes(fill=Cell_Type, y=Count, x=sample_name)) + 
    geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Bar plot of MAMs for all patients") + 
  scale_fill_manual(values=c("#59a89c"))
dev.off()


# T_cells_count

celltype_long <- pivot_longer(celltype, cols = c(T_cells_count), names_to = "Cell_Type", values_to = "Count")
celltype_longtry = celltype_long [,c("sample_name","Cell_Type","Count")]
# Stacked
png ("/omics/groups/OE0436/data/Linh/Output/DEA/T_Count.png",width =700, height = 550)
T = ggplot(celltype_longtry, aes(fill=Cell_Type, y=Count, x=sample_name)) + 
    geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Bar plot of T cells for all patients") +
  scale_fill_manual(values=c("#0b81a2"))
dev.off()


# B_count
celltype_long <- pivot_longer(celltype, cols = c(B_count), names_to = "Cell_Type", values_to = "Count")
celltype_longtry = celltype_long [,c("sample_name","Cell_Type","Count")]
# Stacked
png ("/omics/groups/OE0436/data/Linh/Output/DEA/B_Count.png",width =700, height = 550)
B = ggplot(celltype_longtry, aes(fill=Cell_Type, y=Count, x=sample_name)) + 
    geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Bar plot of B cells for all patients") + scale_fill_manual(values=c("#e25759"))
dev.off()


# DC_count
celltype_long <- pivot_longer(celltype, cols = c(DC_count), names_to = "Cell_Type", values_to = "Count")
celltype_longtry = celltype_long [,c("sample_name","Cell_Type","Count")]
# Stacked
png ("/omics/groups/OE0436/data/Linh/Output/DEA/DC_count.png",width =700, height = 400)
DC = ggplot(celltype_longtry, aes(fill=Cell_Type, y=Count, x=sample_name)) + 
    geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Bar plot of DC count for all patients") +scale_fill_manual(values=c("#7E4794"))
dev.off()

combine_plot = wrap_plots (MP, T,B,DC, ncol =2)
combine_plot
```

```{r}

# MP_count plot
celltype_long <- pivot_longer(celltype, cols = c(MP_count), names_to = "Cell_Type", values_to = "Count")
celltype_longtry <- celltype_long[, c("sample_name", "Cell_Type", "Count")]
MP <- ggplot(celltype_longtry, aes(fill = Cell_Type, y = Count, x = sample_name)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Bar plot of MAMs for all patients") + 
  scale_fill_manual(values = c("#59a89c"))

# T_cells_count plot
celltype_long <- pivot_longer(celltype, cols = c(T_cells_count), names_to = "Cell_Type", values_to = "Count")
celltype_longtry <- celltype_long[, c("sample_name", "Cell_Type", "Count")]
T <- ggplot(celltype_longtry, aes(fill = Cell_Type, y = Count, x = sample_name)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Bar plot of T cells for all patients") +
  scale_fill_manual(values = c("#0b81a2"))

# B_count plot
celltype_long <- pivot_longer(celltype, cols = c(B_count), names_to = "Cell_Type", values_to = "Count")
celltype_longtry <- celltype_long[, c("sample_name", "Cell_Type", "Count")]
B <- ggplot(celltype_longtry, aes(fill = Cell_Type, y = Count, x = sample_name)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Bar plot of B cells for all patients") + 
  scale_fill_manual(values = c("#e25759"))

# DC_count plot
celltype_long <- pivot_longer(celltype, cols = c(DC_count), names_to = "Cell_Type", values_to = "Count")
celltype_longtry <- celltype_long[, c("sample_name", "Cell_Type", "Count")]
DC <- ggplot(celltype_longtry, aes(fill = Cell_Type, y = Count, x = sample_name)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Bar plot of DC count for all patients") + 
  scale_fill_manual(values = c("#7E4794"))

# Combine plots with shared x-axis
combine_plot <- (MP + T) / (B + DC) + plot_layout(guides = "collect")
combine_plot
```

