---
title: "DEA_Classification2"
output: html_document
date: "2024-05-17"
---


The patients will be classified into different groups first using TLS signatures genes and TLS-related cells (MP, T, B, DC) and the present of those cells 


```{r library}

library(Seurat)
#library(leiden)
#library(leidenAlg)
library(ggplot2)
#library(sctransform)
#library(scDblFinder)
#library(SingleCellExperiment)
#library("viridis")
#library(knitr)
library(dplyr)
#library(readxl)
#library(SingleR)
#library(celldex)
library(ggplot2)
library(patchwork)
#library ("harmony")
library (hdf5r)

```



# Loading objects 

GSE_merge was created in DEA_Classification

```{r}
readRDS(GSE_merge,".../Datasets/Merged/GSE_merge_annotated") #GSE 3 datasets,annotated
saveRDS (patient_list, ".../Datasets/DEA/patient_list") #only immune cells, around 25k. Saved 13.05 - run second time
celltype = readRDS( ".../Datasets/DEA/cell_type")

saveRDS(GSE_rest2, ".../Datasets/DEA/GSE_rest2") #only samples that passed the threshold for cell numbers
saveRDS (geometric_means, ".../Datasets/DEA/TLS_score_allcells")

```

# Importing datasets

```{r Import core signatures}
tls_core_signature <- c("CCL2","CCL3","CCL4","CCL5","CCL8","CCL18","CCL19","CCL21","CXCL9","CXCL10","CXCL11","CXCL12", "CXCL13","CD79B", 
"CETP","LAT", "CD1D", "PTGDS","CXCR5","SELL","ICOS","SH2D1A","TIGIT","PDCD1",                 
"BANK1","CD22","CD79A","CR1","FCRL2", "MS4A1", "FCER2", "LAMP3","CD86","CD80","CD83","CCR7")

```

# 1. Level of classification: check for present of B,T,and DC cells

patient_list = each element is a sparse matrix contains normalised data for TLS signature for a patient
patient_list was created in DEA_Classification

celltype: dataframe contains number of cell types for each patients

```{r}
# Calculate the percentage of B_cells and T cells  for each row
celltype$Total_Count <- rowSums(celltype)
celltype$B_Percentage <- celltype$B_count / celltype$Total_Count * 100
celltype$T_Percentage <- celltype$T_cells_count / celltype$Total_Count * 100

# Find the rownames where the B cells percentage is less than 3 %
TLS_rest <- rownames(celltype[celltype$B_Percentage > 1.5 & celltype$T_Percentage > 3 & celltype$DC_count > 0, ]) #10 samples that pass this threshold 

TLSneg_NC = setdiff(rownames(celltype) , TLS_rest) # 23 samples 

# subsetting the samples to calculate TLS score
GSE_list = list()
for (TLS in TLS_rest) {
  GSE_list [[TLS]] = subset (GSE_merge, subset = sample_ID == TLS)  # Warning: Removing 498 cells missing data for vars requested Warning: longer object length is not a multiple of shorter object length
}

GSE_rest2 = merge(GSE_list[[1]],c(GSE_list[[2]],GSE_list[[3]],GSE_list[[4]],GSE_list[[5]],GSE_list[[6]],GSE_list[[7]],GSE_list[[8]],GSE_list[[9]],GSE_list[[10]]))


#rm (GSE_list) # to free memory
# Join layers
GSE_rest2@assays[["RNA"]] = JoinLayers(GSE_rest2@assays[["RNA"]]) #GSE_rest2: 25k cells 
saveRDS(GSE_rest2, ".../Datasets/DEA/GSE_rest2") #only immune cells, around 25k. Saved 13.05 - run second time

```


# 2. Level of classification: check for TLS scores using TLS-related cells

```{r}

# subsetting for a group of cells for classifcation 
cell_sub = c("MAMs:APOE+", "MAMs:S100A8+","T:CD4+:CM1", "T:CD8+:EM","Treg", "T:CD4+:CM2","T:CM", "B-c2","B-c1", "cDC2:CD1C+/CLEC10A+")

GSE_list = list()
GSE_sub = subset (GSE_rest2, subset = ct_ann_GSE344_10kSingleR == "MAMs:APOE+")  #initalize GSE_sub, otherwise wont work
for (i in 2:length(cell_sub)) {
  GSE_list [[i]] = subset (GSE_rest2, subset = ct_ann_GSE344_10kSingleR == cell_sub[i])  # Warning: Removing 143 cells missing data for vars requested Warning: longer object length is not a multiple of shorter object length
  GSE_sub = merge(GSE_sub, GSE_list[[i]])
}

#rm (GSE_list) # to free memory
# Join layers
GSE_sub@assays[["RNA"]] = JoinLayers(GSE_sub@assays[["RNA"]]) #merge layers

# number of cells 
length(Cells(GSE_sub)) #14269 vs 25k cells (50%)

```

## Creating list of patients
 
At the end I get a list of patients, each element is a sparse matrix contains normalised data for TLS signature for a patient

```{r }
patients = unique (GSE_sub@meta.data[["sample_ID"]])
patient_list = list ()

for (i in 1:length(patients)){
  seurat_patient = subset (GSE_sub, subset = sample_ID == patients [i]) 
  matrix_patient = seurat_patient@assays[["RNA"]]@layers[["data"]] # extracting dgcmatrix (matrix patient contain norm data for each patient)
  matrix_patient@Dimnames[[1]] = Features(seurat_patient)  # give genes name
  matrix_patient@Dimnames[[2]] = Cells(seurat_patient) # give cell barcode
  matrix_patient  = matrix_patient [tls_core_signature, ] # subsetting for TLS signatures
  
  patient_list[[patients[i]]] = matrix_patient  #adding to a list
}
#saveRDS (patient_list, ".../Datasets/DEA/patient_list") 
#patient_list = readRDS (".../Datasets/DEA/patient_list")
```

## Calculating mean for each patients

```{r}

for (i in 1:length(patient_list)) {
  mean = rowMeans (patient_list [[i]]) #calculating the mean
  patient_list [[i]] = mean
}
```

## Merge all patients into one dataframe 
```{r}
patient_df = do.call(cbind,lapply(patient_list, as.data.frame)) # do.call(): call a function with a list of arguments. Useful when: have a list of arguments (stored as a list) and want to pass a func to it. lapply () take 2 arguments: the list or vector to apply func to and the func to apply. Short for list apply. lapply and sapply: lapply always return a list
colnames(patient_df) = patients
```


## Calculating TLS score 
calculating the geometric mean value as TLS score

```{r}

library(psych)
patient_df_try = patient_df*1000
patient_df_try = patient_df_try + 1 #to handle 0 values

geometric_means <- apply(patient_df_try, 2, geometric.mean) #2: columns # alot of 0 initially 

geometric_means 
# TLSneg = GSE174401_904 GSE186344_898 GSE174401_915 GSE234832_324 GSE174401_912
# TLSpos = GSE186344_888, 890, 896, GSE186344_902, GSE174401_909
```

```{r}

# extracting the patients ID 
patient_ID = names(geometric_means)
geometric_means = as.data.frame(geometric_means)
geometric_means = cbind (geometric_means, rownames(geometric_means))
colnames(geometric_means) = c("score", "patients")

score_only = geometric_means$score
geometric_means$score = as.numeric(geometric_means$score)
#boxplot (as.numeric(score_only)) 
#points(as.numeric(score_only))

png (".../Output/DEA/TLS_Score_subset2.png",width = 700, height = 1000)
ggplot(geometric_means, aes(x = "Patient_ID",y = as.numeric(score))) + geom_boxplot(fill = "transparent") + geom_point(color="blue", size= 2 , alpha=0.9) + labs(titel="TLS Signature Scores", y = "Scores")
dev.off()

```


# 3. Level of classification: check for TLS scores using all cells

## Creating list of patients
 
At the end I get a list of patients, each element is a sparse matrix contains normalised data for TLS signature for a patient

```{r }
patients = unique (GSE_rest2@meta.data[["sample_ID"]])
patient_list = list ()

for (i in 1:length(patients)){
  seurat_patient = subset (GSE_rest2, subset = sample_ID == patients [i]) 
  matrix_patient = seurat_patient@assays[["RNA"]]@layers[["data"]] # extracting dgcmatrix (matrix patient contain norm data for each patient)
  matrix_patient@Dimnames[[1]] = Features(seurat_patient)  # give genes name
  matrix_patient@Dimnames[[2]] = Cells(seurat_patient) # give cell barcode
  matrix_patient  = matrix_patient [tls_core_signature, ] # subsetting for TLS signatures
  
  patient_list[[patients[i]]] = matrix_patient  #adding to a list
}
#saveRDS (patient_list, ".../Datasets/DEA/patient_list") 
#patient_list = readRDS (".../Datasets/DEA/patient_list")
```

## Calculating mean for each patients

```{r}

for (i in 1:length(patient_list)) {
  mean = rowMeans (patient_list [[i]]) #calculating the mean
  patient_list [[i]] = mean
}
```

## Merge all patients into one dataframe 
```{r}
patient_df = do.call(cbind,lapply(patient_list, as.data.frame)) # do.call(): call a function with a list of arguments. Useful when: have a list of arguments (stored as a list) and want to pass a func to it. lapply () take 2 arguments: the list or vector to apply func to and the func to apply. Short for list apply. lapply and sapply: lapply always return a list
colnames(patient_df) = patients
```


## Calculating TLS score 
calculating the geometric mean value as TLS score

```{r}

library(psych)
patient_df_try = patient_df*1000
patient_df_try = patient_df_try + 1 #to handle 0 values

geo_mean_manually = exp(mean(log(patient_df_try$GSE186344_888))) #geo_mean_manually [1] 109.2226 for 888 -> same value 


geometric_means <- apply(patient_df_try, 2, geometric.mean) #2: columns # alot of 0 initially 

geometric_means 
saveRDS (geometric_means, ".../Datasets/DEA/TLS_score_allcells")
# TLSneg = GSE174401_904 GSE186344_898 GSE174401_915 GSE234832_324 GSE174401_912 (GSE186344_902 changed quite a lot, 20 scores)
```

```{r}

# extracting the patients ID 
patient_ID = names(geometric_means)
geometric_means = as.data.frame(geometric_means)
geometric_means = cbind (geometric_means, rownames(geometric_means))
colnames(geometric_means) = c("score", "patients")

score_only = geometric_means$score
geometric_means$score = as.numeric(geometric_means$score)
#boxplot (as.numeric(score_only)) 
#points(as.numeric(score_only))

png (".../Output/DEA/TLS_Score_2.png",width = 700, height = 1000)
ggplot(geometric_means, aes(x = "Patient_ID",y = as.numeric(score))) + geom_boxplot(fill = "transparent") + geom_point(color="blue", size= 2 , alpha=0.9) + labs(titel="TLS Signature Scores", y = "Scores")
dev.off()

```

