---
title: "ct_ann_final"
output: html_document
date: "2024-05-08"
---

```{r library}

library(tidyverse) # includes ggplot2, for data visualisation. dplyr, for data manipulation.
library(celldex)
library(SingleR)
library(Seurat)
library (ggplot2)
library (VennDiagram)#install.packages ("VennDiagram")
library(eulerr) #install.packages ("eulerr")
set.seed(0)

```

# Annotation with GSE344 taking too long

Problem: too many cells (76k vs 256 in the other ref) in ref dataset -> 3 approaches 

```{r}

BluePrint = as.data.frame(ct_fine@listData[["pruned.labels"]]) # number of MP in GSE344 (ref) 2828, while with blueprint 3474 MP identified -> makes sense

```

#1 Maybe take only the MP + TLS-related cells. What are those? 

```{r}
unique(new_row2)

```

#2 In my 100k cells, 76k from GSE344 already been assigned. Subset the other 2 datasets 

```{r}


```

#3.1 Taking only a subset cells of each cell types

objectives: GSE344: dgCmatrix, new_row: cell type info

```{r}
# adding new_row to GSE344
GSE344@factors = as.vector(new_row)

GSE344_sub = GSE344 [,1:1000] # subsetting 1000 cells 
GSE344_sub@factors =  as.vector(new_row) [1:1000] # adding cell type info
write.table (GSE344_sub,"/omics/groups/OE0436/data/Linh/Datasets/Annotation/ref_GSE344_sub" )

# checking how many cells of interest are in this subset of ref 
new_row_sub = as.data.frame(new_row [1:1000,])
# 193 Macrophages, 100 B cells (B-c1 and B-c2), 110 cells: T:CD4+:CM1,T:CD4+:CM2,T:CD8+:EM, T:CM, 80 cells Dendritic cells: cDC2:CD1C+/CLEC10A+, 16 Treg,.... 

new_row3 = as.data.frame(new_row [1:1000,])

```

#3.2 Taking only a subset cells of each cell types: random

```{r}

GSE_integrated = readRDS("/omics/groups/OE0436/internal/Linh/results/GSE_integrated")

# GSE_integrated subset for GSE344: 

GSE344 = subset(GSE_integrated, subset = dataset_ID == "GSE186344")

new_row2 = read.table ("/omics/groups/OE0436/data/Linh/Datasets/GSE186344/new_row2")
GSE344 = AddMetaData(GSE344, as.vector(new_row2)) #adding cell type info
GSE344 = AddMetaData(GSE344, Cells(GSE344), col.name = "cell_barcode")
# taking a subset of cells randomly, using subset and sample_int
set.seed(0) #so that its reproducible 
GSE344_matrix = GSE344@assays[["SCT"]]@data # extract the normalised count as well as genes and cell barcode
random_cells = sample(GSE344_matrix@Dimnames[[2]],10000)

GSE344_10k <- GSE344[,random_cells] # GSE344 automatically order it after subsetting 

#generate five random numbers between 1 and 20
saveRDS (GSE344_10k,"/omics/groups/OE0436/data/Linh/Datasets/Annotation/ct_ann_GSE344_10k")

```

```{r}
# Checking cell type distribution 

GSE344_ran_value = GSE344_ran@assays[["SCT"]]@data # extract the dgcmatrix 
 ()
celltype = as.data.frame(GSE344_ran@meta.data[["x"]]) #extracting the cell type 
#cell distribtion: 82 Macrophages, 43 B cells, 175 + 76 T cells, 26 DCs 

```


# Result from 3.1

```{r result from #3.1 approach}

ct_ann_GSE344sub = readRDS ("/omics/groups/OE0436/data/Linh/Datasets/Annotation/ct_ann_GSE344sub")
GSE_integrated@meta.data[["ct_ann_GSE344sub"]] = ct_ann_GSE344sub@listData[["pruned.labels"]] #adding the metadata info 

GSE_integrated <- SetIdent(GSE_integrated, value = "ct_ann_GSE344sub")

# plot UMAP 
png ("/omics/groups/OE0436/data/Linh/Output/Annotation/GSE186344/UMAP_Anno",width = 1000, height = 700) #res =?
DimPlot(GSE_integrated, label = T , repel = T, label.size = 3) + NoLegend()
dev.off()
# plot Heatmap
png ("/omics/groups/OE0436/data/Linh/Output/Annotation/GSE186344/Heatmap",width = 1000, height = 700)
plotScoreHeatmap(ct_ann_GSE344sub)
dev.off()
# plot Delta Distribution
png ("/omics/groups/OE0436/data/Linh/Output/Annotation/GSE186344/DeltaDistribution",width = 1000, height = 700)
plotDeltaDistribution(ct_ann_GSE344sub, ncol = 4, dots.on.top = FALSE)
dev.off()


```


```{r result from #3.2 approach}

#lost while trying to save the session 

ct_ann_GSE344ran = readRDS ("/omics/groups/OE0436/data/Linh/Datasets/Annotation/ct_ann_GSE344ran")
GSE_integrated@meta.data[["ct_ann_GSE344ran"]] = ct_ann_GSE344ran@listData[["pruned.labels"]] #adding the metadata info 


```

```{r result from #3.3 approach}


ct_ann_GSE344_10kSingleR = readRDS ("/omics/groups/OE0436/data/Linh/Datasets/Annotation/ct_ann_GSE344_10kSingleR")
GSE_integrated@meta.data[["ct_ann_GSE344_10kSingleR"]] = ct_ann_GSE344_10kSingleR@listData[["pruned.labels"]] #adding the metadata info 

GSE_integrated <- SetIdent(GSE_integrated, value = "ct_ann_GSE344_10kSingleR")

# plot UMAP 
png ("/omics/groups/OE0436/data/Linh/Output/Annotation/GSE186344/10k_random/UMAP_Anno",width = 1000, height = 700) #res =?
DimPlot(GSE_integrated, label = T , repel = T, label.size = 3) + NoLegend()
dev.off()
# plot Heatmap
png ("/omics/groups/OE0436/data/Linh/Output/Annotation/GSE186344/10k_random/Heatmap",width = 1000, height = 700)
plotScoreHeatmap(ct_ann_GSE344_10kSingleR)
dev.off()
# plot Delta Distribution
png ("/omics/groups/OE0436/data/Linh/Output/Annotation/GSE186344/10k_random/DeltaDistribution",width = 1000, height = 700)
plotDeltaDistribution(ct_ann_GSE344_10kSingleR, ncol = 4, dots.on.top = FALSE)
dev.off()



```

# Venn Diagramm

#1 Should GSE344 be assigned again 

```{r }

# GSE344_celltype that were GIVEN 
new_row2 = read.table ("/omics/groups/OE0436/data/Linh/Datasets/GSE186344/new_row2")



```

```{r}
# GSE344_celltype that were ASSIGNED
# GSE_integrated subset for GSE344: 


# 2000 random 
ct_ann_GSE344ran = readRDS ("/omics/groups/OE0436/data/Linh/Datasets/Annotation/ct_ann_GSE344ran")
GSE_integrated = readRDS("/omics/groups/OE0436/internal/Linh/results/GSE_integrated")
GSE_integrated@meta.data[["ct_ann_GSE344ran"]] = ct_ann_GSE344ran@listData[["pruned.labels"]] #adding the metadata info 
GSE344 = subset(GSE_integrated, subset = dataset_ID == "GSE186344")
GSE344_assigned = GSE344@meta.data[["ct_ann_GSE344ran"]]


# 1000 non random 
ct_ann_GSE344sub = readRDS ("/omics/groups/OE0436/data/Linh/Datasets/Annotation/ct_ann_GSE344sub")
GSE_integrated@meta.data[["ct_ann_GSE344sub"]] = ct_ann_GSE344sub@listData[["pruned.labels"]] #adding the metadata info 
GSE344 = subset(GSE_integrated, subset = dataset_ID == "GSE186344")
GSE344_assigned_sub = GSE344@meta.data[["ct_ann_GSE344sub"]]

# 10000 random
ct_ann_GSE344_10kSingleR = readRDS ("/omics/groups/OE0436/data/Linh/Datasets/Annotation/ct_ann_GSE344_10kSingleR")
GSE_integrated@meta.data[["ct_ann_GSE344_10kSingleR"]] = ct_ann_GSE344_10kSingleR@listData[["pruned.labels"]] #adding the metadata info 
GSE344 = subset(GSE_integrated, subset = dataset_ID == "GSE186344")
GSE344_assigned_10k = GSE344@meta.data[["ct_ann_GSE344_10kSingleR"]]


```

```{r Pie chart}

########################### 2000 random 
overlap_GSE344 = (new_row2[,1]== GSE344_assigned) #creating a logical vector
match = sum (overlap_GSE344, na.rm = TRUE)# number of matching elements #53927
non_match_count = sum (overlap_GSE344 == F, na.rm = TRUE) #11920
NA_count = sum (is.na(overlap_GSE344)) #10824: the number of cells that cell type were not GIVEN. The new_row are NA itself 
#creating a pie chart

#percentage = c(match,non_match_count,NA_count)
#percentage <- round(percentage/76671*100)
#percentage <- paste(percentage,"%",sep="") # ad % to labels
png ("/omics/groups/OE0436/data/Linh/Output/Annotation/GSE186344/GSE186344 cell type given vs. assigned (2k random cells)",width = 1000, height = 700)
pie(c(match,non_match_count,NA_count), labels = c("Matching", "Non-Matching", "NA"), col= c("skyblue", "lightcoral", "lightgreen"),main = "GSE186344 cell type given vs. assigned (2k random cells)") 
dev.off()

# 1000 non random
overlap_GSE344_sub =  (new_row2[,1]== GSE344_assigned_sub)
match = sum (overlap_GSE344_sub == "TRUE", na.rm = TRUE) #36884
non_match_count = sum (overlap_GSE344_sub == "FALSE", na.rm = TRUE) #28994
NA_count = sum (is.na(overlap_GSE344_sub)) #10793: the number of cells that cell type were not GIVEN


png ("/omics/groups/OE0436/data/Linh/Output/Annotation/GSE186344/GSE186344 cell type given vs. assigned (1k cells)",width = 1000, height = 700)
pie(c(match,non_match_count,NA_count), labels = c("Matching", "Non-Matching", "NA"), col= c("skyblue", "lightcoral", "lightgreen"),main = "GSE186344 cell type given vs. assigned (1k cells)") 
dev.off()

# 10000 random
overlap_GSE344 = (new_row2[,1] == GSE344_assigned_10k) #creating a logical vector 
match = sum(overlap_GSE344, na.rm = TRUE)# number of matching elements 54255
non_match_count = sum (overlap_GSE344 == F, na.rm = TRUE)#11423
NA_count = sum (is.na(overlap_GSE344)) #10993: the number of cells that cell type were not GIVEN. The new_row are NA itself 

png ("/omics/groups/OE0436/data/Linh/Output/Annotation/GSE186344/GSE186344 cell type given vs. assigned (10k cells).png",width = 1000, height = 700)
pie(c(match,non_match_count,NA_count), labels = c("Matching", "Non-Matching", "NA"), col= c("skyblue", "lightcoral", "lightgreen"),main = "GSE186344 cell type given vs. assigned (10k cells)") 
dev.off()

```
# 2 Comparing 3 ref datasets from GSE344 and the given annotation

```{r}
# define 4 vectors: 1. given: new_row2 2. GSE344_assigned_10k 3.GSE344_assigned_sub (1k) 4. GSE344_assigned (2k)
one = new_row2
two = GSE344_assigned_10k 
three = GSE344_assigned_sub
four = GSE344_assigned
# calculating area of overlaps

n12 = (sum ((one== two)== T, na.rm = T)/ 76671)*100 # length of vector -> get %
n23 = (sum ((two== three)== T, na.rm = T)/ 76671)*100
n13 = (sum ((one== three)== T, na.rm = T)/ 76671)*100
n14 = (sum ((one== four)== T, na.rm = T)/ 76671)*100
n24 = (sum ((two== four)== T, na.rm = T)/ 76671)*100
n34 = (sum ((three== four)== T, na.rm = T)/ 76671)*100
n123 = (length (which(one== two & two == three))/ 76671)*100
n124 = (length (which(one== two & two == four))/ 76671)*100
n234 = (length (which(two== three & three == four))/ 76671)*100
n134 = (length (which(one== three & three == four))/ 76671)*100
n1234 = (length (which(one== two & two== three & three == four)))*100/ 76671

n12 = as.integer (n12)
n23 = as.integer (n23)
n13 = as.integer (n13)
n14 = as.integer (n14)
n24 = as.integer (n24)
n34 = as.integer (n34)
n123 = as.integer (n123)
n124 = as.integer (n124)
n134 = as.integer (n134)
n234 = as.integer (n134)
n1234 = as.integer (n1234)

png ("/omics/groups/OE0436/data/Linh/Output/Annotation/GSE186344/Venn_4refGSE344",width = 1000, height = 700) #res =?
grid.newpage()
draw.quad.venn (area1 = 100, area2 = 100, area3  = 100, area4 = 100, n12 = n12, n13 = n13, n23 = n23, n14 = n14, n24 = n24,n34 =n34,  n123 =n123, n124 =n124,n234 =n234, n134 =n134, n1234= n1234, category = c("given","1k_nonrandom","2k", "10k"), fill =c("skyblue", "lightcoral", "lightgreen","lightyellow"))
dev.off()

# using euler

#specify values to use in venn diagram
fit <- euler(c(area1 = 100, area2 = 100, area3  = 100, area4 = 100, n12 = n12, n13 = n13, n23 = n23, n14 = n14, n24 = n24,n34 =n34,  n123 =n123, n124 =n124,n234 =n234, n134 =n134, n1234= n1234))

#create venn diagram with custom colors
plot(fit, fill=c("skyblue", "lightcoral", "lightgreen","lightyellow")) 
```



# Comparing 4 ref datasets

- Compare for all cells: not very feasible since cell type names are different
=> Compare for MP,B,T,DC

```{r Importing the cell type infos }
ct_ann_THCA  = readRDS ("/omics/groups/OE0436/data/Linh/Datasets/Annotation/ct_ann.rds")
ct_ann_main_BluePrint = readRDS ("/omics/groups/OE0436/data/Linh/Datasets/Annotation/ct_main_BluePrint")
ct_ann_GSE344_10kSingleR  =  readRDS ("/omics/groups/OE0436/data/Linh/Datasets/Annotation/ct_ann_GSE344_10kSingleR")
```


```{r Adding cell type annotation to GSE_integrated}

GSE_integrated@meta.data[["ct_ann_GSE344_10kSingleR"]] = ct_ann_GSE344_10kSingleR@listData[["pruned.labels"]] #adding the metadata info 
GSE_integrated@meta.data[["ct_ann_main_BluePrint"]] = ct_ann_main_BluePrint@listData[["pruned.labels"]] #adding the metadata info 
GSE_integrated@meta.data[["ct_ann_THCA"]] = ct_ann_THCA[,"pruned.labels"] #adding the metadata info 


#saveRDS (GSE_integrated)
```

```{r Susetting for cell type (MP, T, B, DC)}

# for BluePrint: "Macrophages", c("CD4+ T-cells", "CD8+ T-cells"), "B-cells", "DC"
# for TGCA: "Macrophage", "T_cells", "B_cell", "DC"
# for MP: contain c("MAMs:APOE+", "MAMs:S100A8+"),c("T:CD4+:CM1", "T:CD8+:EM","Treg", "T:CD4+:CM2","T:CM"),c("B-c2","B-c1"),"cDC2:CD1C+/CLEC10A+"
# for tumor cells:  "MTCs"     

# define the names of the cells 
BluePrint_MP =  "Macrophages"
BluePrint_T =  c("CD4+ T-cells", "CD8+ T-cells")
BluePrint_B =  "B-cells"
BluePrint_DC = "DC"

GSE344_MP = c("MAMs:APOE+", "MAMs:S100A8+")
GSE344_T =  c("T:CD4+:CM1", "T:CD8+:EM","Treg", "T:CD4+:CM2","T:CM")
GSE344_B =  c("B-c2","B-c1")
GSE344_DC = "cDC2:CD1C+/CLEC10A+"

TGCA_MP ="Macrophage" 
TGCA_T =  "T_cells"
TGCA_B =  "B_cell"
TGCA_DC = "DC"

# subsetting the according cell populations
BluePrint_MP =  subset (GSE_integrated, subset = ct_ann_main_BluePrint == BluePrint_MP)
BluePrint_T = subset (GSE_integrated, subset = ct_ann_main_BluePrint == BluePrint_T)
BluePrint_B =  subset (GSE_integrated, subset = ct_ann_main_BluePrint == BluePrint_B)
BluePrint_DC = subset (GSE_integrated, subset = ct_ann_main_BluePrint == BluePrint_DC)

GSE344_MP = subset (GSE_integrated, subset = ct_ann_GSE344_10kSingleR == GSE344_MP)
GSE344_T =  subset (GSE_integrated, subset = ct_ann_GSE344_10kSingleR == GSE344_T)
GSE344_B =  subset (GSE_integrated, subset = ct_ann_GSE344_10kSingleR == GSE344_B)
GSE344_DC = subset (GSE_integrated, subset = ct_ann_GSE344_10kSingleR == GSE344_DC)

TGCA_MP = subset (GSE_integrated, subset = ct_ann_THCA == TGCA_MP)
TGCA_T =  subset (GSE_integrated, subset = ct_ann_THCA == TGCA_T)
TGCA_B =  subset (GSE_integrated, subset = ct_ann_THCA == TGCA_B)
TGCA_DC = subset (GSE_integrated, subset = ct_ann_THCA == TGCA_DC)

```

```{r Extract cell barcode}
BluePrint_MP =  Cells (BluePrint_MP)
BluePrint_T =   Cells (BluePrint_T)
BluePrint_B =   Cells (BluePrint_B)
BluePrint_DC =  Cells (BluePrint_DC)

GSE344_MP = Cells (GSE344_MP)
GSE344_T =   Cells (GSE344_T)
GSE344_B =   Cells (GSE344_B)
GSE344_DC =  Cells (GSE344_DC)

TGCA_MP =  Cells (TGCA_MP)
TGCA_T =   Cells (TGCA_T)
TGCA_B =   Cells (TGCA_B)
TGCA_DC =  Cells (TGCA_DC)
```

```{r Venn diagram for overlap}
 ############################ MP 
#specify values to use in venn diagram
fit <- euler(c('BluePrint_MP' = length(BluePrint_MP), 'GSE344_MP' = length(GSE344_MP), "TGCA_MP" = length(TGCA_MP), 'BluePrint_MP&GSE344_MP' = length(intersect(BluePrint_MP, GSE344_MP)), 'GSE344_MP&TGCA_MP' = length(intersect(GSE344_MP, TGCA_MP)),'BluePrint_MP&TGCA_MP' = length(intersect(BluePrint_MP,TGCA_MP)), 'BluePrint_MP&GSE344_MP&TGCA_MP' =  length(intersect(intersect(BluePrint_MP,GSE344_MP),TGCA_MP))))
#create venn diagram with custom colors
png ("/omics/groups/OE0436/data/Linh/Output/Annotation/GSE186344/MP_Overlap",width = 1000, height = 700) #res =?
plot(fit, fill=c("skyblue", "lightcoral", "lightgreen"),quantities = TRUE, shape = "ellipse") 
dev.off()


 ############################ T 
fit <- euler(c('BluePrint_T' = length(BluePrint_T), 'GSE344_T' = length(GSE344_T), "TGCA_T" = length(TGCA_T), 'BluePrint_T&GSE344_T' = length(intersect(BluePrint_T, GSE344_T)), 'GSE344_T&TGCA_T' = length(intersect(GSE344_T, TGCA_T)),'BluePrint_T&TGCA_T' = length(intersect(BluePrint_T,TGCA_T)), 'BluePrint_T&GSE344_T&TGCA_T' = length(intersect(intersect(BluePrint_T,GSE344_T),TGCA_T))))
#create venn diagram with custom colors
png ("/omics/groups/OE0436/data/Linh/Output/Annotation/GSE186344/T_Overlap",width = 1000, height = 700) #res =?
plot(fit, fill=c("skyblue", "lightcoral", "lightgreen"),quantities = TRUE, shape = "ellipse") 
dev.off()

 ############################ B 
fit <- euler(c('BluePrint_B' = length(BluePrint_B), 'GSE344_B' = length(GSE344_B), "TGCA_B" = length(TGCA_B), 'BluePrint_B&GSE344_B' = length(intersect(BluePrint_B, GSE344_B)), 'GSE344_B&TGCA_B' = length(intersect(GSE344_B, TGCA_B)),'BluePrint_B&TGCA_B' = length(intersect(BluePrint_B,TGCA_B)), 'BluePrint_B&GSE344_B&TGCA_B' = length(intersect(intersect(BluePrint_B,GSE344_B),TGCA_B))))
#create venn diagram with custom colors
png ("/omics/groups/OE0436/data/Linh/Output/Annotation/GSE186344/B_Overlap",width = 1000, height = 700) #res =?
plot(fit, fill=c("skyblue", "lightcoral", "lightgreen"),quantities = TRUE, shape = "ellipse") 
dev.off()


 ############################ DC
fit <- euler(c('BluePrint_DC' = length(BluePrint_DC), 'GSE344_DC' = length(GSE344_DC), "TGCA_DC" = length(TGCA_DC), 'BluePrint_DC&GSE344_DC' = length(intersect(BluePrint_DC, GSE344_DC)), 'GSE344_DC&TGCA_DC' = length(intersect(GSE344_DC, TGCA_DC)),'BluePrint_DC&TGCA_DC' = length(intersect(BluePrint_DC,TGCA_DC)), 'BluePrint_DC&GSE344_DC&TGCA_DC' = length(intersect(intersect(BluePrint_DC,GSE344_DC),TGCA_DC))))
#create venn diagram with custom colors
png ("/omics/groups/OE0436/data/Linh/Output/Annotation/GSE186344/DC_Overlap",width = 1000, height = 700) #res =?
plot(fit, fill=c("skyblue", "lightcoral", "lightgreen"),quantities = TRUE, shape = "ellipse") 
dev.off()

```

